The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R

### 2023-11-23 manuscript/manuscript.qmd
final work through manuscript and minor edits

Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved and hence can be seen as a "*time machine*" that enables researchers and developers to compare code and repositories across different versions over time. For our purposes, a repository can be seen as an isolated project directory which includes all necessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).

@@

In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies. For the following examples, we focus on workflows utilizing R/the RStudio integrated development environment (IDE) GUI ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively. We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI. For more details, we refer to the book *Pro Git* by Chacon and Straub.[@chacon2014pro]

@@

A special aspect about real-world data, such as administrative insurance claims and electronic health records, is that these data are typically not collected for the primary purposes of doing research but rather for reasons of billing or clinical documentation. Hence, being transparent about the data provenance, pre-processing steps and operationalization of exposures, outcomes and covariates is of critical importance [@wang2017] and should be part of the repository's metadata. This can come in form of codebooks, data dictionaries, literate programming scripts (e.g., Quarto) or other applicable documentation files. Applying and documenting naming conventions of files, data structures and variables as developing reproducible and well documented functions or packages as well as inline commenting of programming code can further enhance the transparency and readability of analytic code.[@bové2023improving]

@@

Every time a user makes a commit to create a snapshot of the work, `Git` creates a unique hash (a 40-character string created by a **simple hashing algorithm** \[SHA\]) which can be seen as an "ID" of the snapshot which enables users to comprehend every change made and revert back to any snapshot of the repository. @fig-stage-commit-RStudio-gui illustrates the equivalent steps in RStudio GUI.

@@

#### Tags (`tag`) {#sec-tags}

A handy `Git` feature to highlight particularly important commits, snapshots, code/software releases or milestones of a project is `tagging`. In context of RWE studies, this can be a specific version of an analysis, e.g., the one used for regulatory submission or completion of a manuscript version or revision.

@@

```{bash}
git push origin v1.0
```

@@

### Synchronize remote to local (`pull`)

In case collaborators have also made modification to their local files and pushed them to the remote repository, these changes can be easily synchronized with our local copy via the `pull` command.

```{bash}
# pull remote repository version of code to local
git pull
```

@@

```{bash}
# Example .gitgnore file specyfing which 
# files, sub-directories and patterns
# should not be tracked

@@

# .Renviron in root directory
.Renviron
```

@@

```{r}
# Example how the credentials defined in .Renviron can be used
# inside an RScript
set_synapse_credentials(
  username = Sys.getenv("db_username"), 
  password = Sys.getenv("db_password")
  )
```

Third, it needs to be ensured that access permissions to remote repositories are appropriately configured, granting only necessary access levels to collaborators unless it is intended to share the repository with the public. All major remote repositories offer the option to configure a repository to be private or public.

@@

Besides the usage of `Git` and remote repositories as a version control systems, there are also further relevant aspects that determine the reproducibility of analytic code when using open-source software such as R. To complement the introduction and discussion on `Git` in this manuscript, section 2 (*Reproducible computation environments in R*) in the **Supplementary Material** provides a few further considerations addressing the computational reproducibility when using R.

@@

Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms. `Git` was primarily designed for version control and development of analytic code by statisticians and programmers, but may also benefit other members of a RWE generation team through the centralized deployment, tracking and version control of study documents as well as project management tools which have become standard among most established remote repository platforms. While `Git` offers substantial benefits for collaboration, reproducibility, and transparency in RWE teams, challenges can arise in its adoption, particularly related to merging conflicts, steep learning curves, and maintaining consistent usage practices across team members.

@@

```{r}
#| label: fig-RStudio-gui
#| fig-cap: "Steps to clone a remote repository using the RStudio graphical user interface."
#| eval: true
#| echo: false
#| out-width: 50%
knitr::include_graphics(here::here("figures", "Figure_1_RStudio_init.png"))
```

@@

```{r}
#| label: fig-stage-commit-RStudio-gui
#| fig-cap: "RStudio provides a graphical user interface as an alternative to command line prompts to interact with Git and to perform command equivalent to git add, git commit and git push/pull (red circles). This view can be accessed in the right upper pane of RStudio under Git > Diff. In this example, an R script with the title '03_propensity_score_analysis.R' is selected and staged. All changes performed in this file can be viewed in the window below with green highlighted lines indicating additions and red highlighted lines deletions from the prior version. In this case, a title was added to a propensity score plot which is also expressed in the commit message."
#| eval: true
#| echo: false
knitr::include_graphics(here::here("figures", "Figure_4_stage_commit.png"))
```

>>>>

Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved and hence can be seen as a "*time machine*" that enables researchers and developers to compare code and repositories across different versions over time. For our purposes, a repository can be thought of as an isolated project directory which includes all necessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).

@@

In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies. For the following examples, we focus on workflows utilizing the RStudio integrated development environment (IDE) GUI ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository platforms, respectively. We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI. For more details, we refer to the book *Pro Git* by Chacon and Straub.[@chacon2014pro]

@@

A special aspect about real-world data, such as administrative insurance claims and electronic health records, is that these data are typically not collected for the primary purposes of doing research but rather for reasons of billing or clinical documentation. Hence, being transparent about the data provenance, pre-processing steps and operationalization of exposures, outcomes and covariates is of critical importance [@wang2017] and should be part of the repository's metadata. This can come in form of codebooks, data dictionaries, literate programming scripts (e.g., Quarto) or other applicable documentation files. Developing reproducible and well documented functions or packages as well as inline commenting of programming code can further enhance the transparency and readability of analytic code.[@bové2023improving]

@@

Every time a user makes a commit to create a snapshot of the work, `Git` creates a unique hash (a 40-character string created by a **simple hashing algorithm** \[SHA\]) which can be seen as an "ID" of the snapshot which enables users to comprehend every change made and revert back to any snapshot of the repository. @fig-stage-commit-RStudio-gui illustrates the equivalent steps in the RStudio GUI.

@@

### Tags (`tag`) {#sec-tags}

A handy `Git` feature to highlight particularly important commits, snapshots, code/software releases or milestones of a project is `tagging`. In context of RWE studies, this can be a specific version of an analysis, e.g., the one used for a regulatory submission or the completion of a manuscript version or revision.

@@

```{bash}
git push origin v2.0
```

@@

@@

```{bash}
# Example .gitgnore file specyfing which 
# files, sub-directories and patterns
# that should not be tracked

@@

# .Renviron file
.Renviron
```

@@

```{r}
# Example how the credentials defined in .Renviron can be used
# inside an R script
set_synapse_credentials(
  username = Sys.getenv("db_username"), 
  password = Sys.getenv("db_password")
  )
```

Third, it needs to be ensured that access permissions to remote repository platforms are appropriately configured, granting only necessary access levels to collaborators unless it is intended to share the repository with the public. All major remote repositories offer the option to configure a repository to be private or public.

@@

Besides the usage of `Git` and remote repositories as version control systems, there are also further relevant aspects that determine the reproducibility of analytic code when using open-source software such as R. To complement the introduction and discussion on `Git` in this manuscript, section 2 (*Reproducible computation environments in R*) in the **Supplementary Material** provides a few further considerations addressing the computational reproducibility when using R.

@@

Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms. `Git` was primarily designed for version control and development of analytic code by statisticians and programmers, but may also benefit other members of a RWE generation team through the centralized deployment, tracking and version control of study documents as well as project management tools which have become standard among established remote repository platforms. While `Git` offers substantial benefits for collaboration, reproducibility, and transparency in RWE teams, challenges can arise in its adoption, particularly related to merging conflicts, steep learning curves, and maintaining consistent usage practices across team members.

@@

```{r}
#| label: fig-RStudio-gui
#| fig-cap: "Steps to clone a remote repository using the RStudio integrated development environment (IDE) graphical user interface (GUI)."
#| eval: true
#| echo: false
#| out-width: 50%
knitr::include_graphics(here::here("figures", "Figure_1_RStudio_init.png"))
```

@@

```{r}
#| label: fig-stage-commit-RStudio-gui
#| fig-cap: "The RStudio integrated development environment (IDE) provides a graphical user interface (GUI) as an alternative to command line prompts to interact with Git and to perform command equivalent to git add, git commit and git push/pull (red circles). This view can be accessed in the right upper pane of the RStudio IDE under Git > Diff. In this example, an R script with the title '03_propensity_score_analysis.R' is selected and staged. All changes performed in this file can be viewed in the window below with green highlighted lines indicating additions and red highlighted lines deletions from the prior version. In this example, a title was added to a propensity score plot which is also expressed in the commit message."
#| eval: true
#| echo: false
knitr::include_graphics(here::here("figures", "Figure_4_stage_commit.png"))
```
